Synthesising Complex Termination Arguments

### 2014-06-23 papers/termination/paper.tex
revised motivation

\section{Motivational Examples} \label{sec:motivation}
We next discus some intricate examples that prove challenging for existent termination analysis.
%a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
%frequently in practise and are not properly handled by existing techniques.

%We now present a few examples that justify

\subsection{Program linearity vs. the linearity of its ranking function.}
The first examples we consider illustrate the lack of a direct correlation between the linearity of a program and that of its termination arguments.
As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
although a linear ranking function does exist (see Figure~\ref{fig:handletable}).

In order to find a ranking function for this example, it is necessary to take into account
the semantics of the bit-wise AND operator, which is not easily done when working with mathematical integers \cite{}.
Our technique finds that a possible ranking function is the linear function
$R(x) = x$, whose value decreases with
every iteration, but it can not decrease indefinitely as it is bounded from below.

Conversely, Figure~\ref{} shows a linear program with a non-linear ranking function.

Figure~\ref{} is a linear program taken from \cite{DBLP:conf/tacas/CookSZ13}, where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a disjunctively well-founded one, 
which requires an expensive binary reachability analysis.
However, by using our technique we can find a non-linear lexicographic ranking functions.
A lexicographic non-linear ranking functions consists of lexicographically ordered components
of non-linear functions. As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
loop transition leads to a decrease with respect to the lexicographic
ordering for this tuple. Therefore no function may increase unless a function of
a lower index decreases. Additionally, at every step, there must be at least one
function that decreases.

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
  while (x != 0) {
    if (x > 0)
      x--;
    else
      x++;
  }
\end{lstlisting}
\caption{}
 \label{fig:motivation.c}
\end{subfigure}%

\end{tabular}
\end{figure*}

\subsection{Differences in the termination behaviour for integers and bit-vectors.}
We have collected a number of motivational examples from other termination papers that treat bit-vectors as mathematical integers \cite{DBLP:conf/tacas/LeikeH14,DBLP:conf/tacas/CookSZ13}. 
We show that the termination arguments computed by such techniques do not directly apply when considering the bit-vector semantics.
%these programs actually exhibit different terminating behaviours for bit-vectors and integer

\begin{lstlisting}
  while (x != m) {
    if (x > m)
      x = 0;
    else
      x++;
  }
\end{lstlisting}

The program in Figure~\ref{} is taken from \cite{DBLP:conf/tacas/CookSZ13}, where
$m$ and $x$ start as any integers with $m$ positive. If $x$ is greater than
$m$, $x$ is set to 0. Otherwise, $x$ increases until it equals $m$, upon which
the loop terminates. While a disjunctive well-founded termination argument does exist for the loop, 
e.g. ($x < oldx$ and $0 \leq oldx$) or ($m-x < oldm-oldx$ and $0 \leq oldm-oldx$), the loop does not 
terminate under the bit-vector semantics. The reason is ..

\begin{lstlisting}
  while ( q >= 0 ) {
    q = q - y;
    y = y + 1;
  }
\end{lstlisting}

In Figure~\ref{}, every execution of the program can be partitioned into two phases: in the first phase $y$ increases
until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
This example was presented in \cite{DBLP:conf/tacas/LeikeH14}, where the authors make use of a template for obtaining a ranking function that proceeds
through a fixed number of phases in the program execution. Each phase is ranked by a linear function, and ends when this function becomes non-positive.
However, when considering the bit-vector semantics, this program does not terminate as..

\subsection{Differences in the termination behaviour for reals and floats.}
\begin{lstlisting}
while (x > 0.0) x -= 1.0;
\end{lstlisting}

\begin{lstlisting}
while (x > 0.0) x *= 0.5;
\end{lstlisting}

\subsection{Multi-phase ranking functions}
%\subsection{Lexicographic ranking function with strict inequalities.}
%Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
The program in Figure~\ref{} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multi-phase ranking function, i.e. ..,  with a dedicated technique for computing it. However, in our setting 
this type of programs do not need a special treatment, as we can find a non-linear ranking function for it as follows:
\begin{verbatim}
R(x, y, z) = (x < y, z)
\end{verbatim}

\begin{lstlisting}
int main(void) {
  int x, y, z;

  while (x > 0 && y > 0 && z > 0) {
    if (y > x) {
      y = z;
      x = nondet();
      z = x - 1;
    } else {
      z = z - 1;
      x = nondet();
      y = x - 1;
    }
  }
}
\end{lstlisting}

@@

As well as machine arithmetic differing from Peano arithmetic on the operators they have in common,
computers have several ``bitwise'' operations that are not taken as primitive in the theory of
integers.  These operations include the standard boolean operations \texttt{and, or, not, xor}
applied to each element of the bit-vector.  Computer programs often make use of these operators,
which are non-linear when interpreted in the standard model of Peano arithmetic\footnote{
Although some of these operators can be seen as linear in a different algebraic structure,
e.g. \texttt{xor} corresponds to addition in the Galois field $\mathrm{GF}(2^k)$}.

@@

\section{Termination as Second-Order Satisfaction}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  In fact, many of these problems can be encoded in a fragment of second-order logic we will
suggestively refer to as the \emph{synthesis fragment}:

\begin{definition}[Synthesis Fragment]
 A formula is in the synthesis fragment iff it is of the form
 \[
  \exists P, x \cdot \forall y \cdot \sigma(P, x, y)
 \]

 Here $P$ ranges over sets, while $x$ and $y$ range over ground terms.
\end{definition}

We will then show that checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis.
We will make use of this observation to frame the halting problem as a program synthesis problem, allowing us
to employ the program synthesiser described in Section~\ref{sec:synthesis}.

\subsection{Variations of the Halting Problem}

@@

\begin{definition}[Conditional Termination Formula]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, V \cdot \forall x, x' \cdot & I(x) \wedge L(x) \rightarrow V(x) ~ \wedge \\
                                 & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Multiple Loops Termination Formula]
\label{def:multi-termination-formula}
 \begin{align*}
  \exists R, V_1, V_2 \cdot \forall x, x' \cdot & I(x) \rightarrow V_1(x) ~ \wedge \\
                                        & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
                                        & V_1(x) \wedge \lnot L_1(x) \wedge L_2(x) \rightarrow V_2(x) ~ \wedge \\
                                        & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

A slightly more involved case occurs if $L$'s termination depends on the state it begins
executing in, i.e. $L$ is \emph{conditionally terminating}.  This property is equivalent to the
formula of Defintion~\ref{def:conditional-termination-formula}.  If this formula is satisfiable,
$V$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
condition is met.
$R$ is then a ranking function for $L$ as restricted by $V$ -- that is to say, $R$ need only
be well founded on those states satisfying $V$.  But since $V$ is an inductive invariant of $L$,
$R$ is strong enough to show that $L$ terminates from any of its initial states.  We say
that $V$ is a \emph{termination invariant} for $L$ and that $R$ proves termination relative to $V$.

@@

\subsection{Program Synthesis as Second-Order Satisfaction}

With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae above
belongs to the synthesis fragment.  We use this name because the program synthesis problem can be
described as finding a satisfying assigment to the following formula, which is the general case of
the synthesis fragment:

\begin{definition}[Synthesis Formula]
 \label{def:synth-formula}
 $$\exists P, x \cdot \forall y \cdot \sigma(P, x, y)$$
\end{definition}

Here we can view $\sigma: X \times Y \times (X \times Y \to Z) \to \mathbb{B}$ as a specification function.
The idea is that $\sigma$ returns true iff the function $P$ computes appropriate outputs
when fed the inputs $x$ and $y$.  If we can find a function $P$ and a value $x$ that meet the specification
for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.

\subsection{Time Complexity}
In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
This procedure has the property that it will find the shortest program meeting the
specification, and as a consequence the runtime of the program synthesiser is dominated
by the size of this shortest program.  To frame the discussion of the runtime of our
termination proving procedure, we first recall the definition of the Kolmogorov complexity
of a function $f$:

>>>>

\section{Motivational Examples} \label{sec:motivation}
In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}). 
Now, we provide some simple examples that break those assumptions and cannot therefore be handled by state of the art tools.

%We next discus some intricate examples that prove challenging for existent termination analysis.

%\subsection{Program linearity vs. the linearity of its ranking function.}

Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.
A program not falling in this category is given in Figure~\ref{fig:motivation.a}.
%As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
%although a linear ranking function does exist (see Figure~\ref{fig:handletable}). 
In order to find a ranking function for this example, it is necessary to take into account
the semantics of the bit-wise AND operator, which is not easily done when working with mathematical integers.
Our technique finds that a possible ranking function is the linear function
$R(x) = x$, whose value decreases with
every iteration, but it can not decrease indefinitely as it is bounded from below.
This example also illustrates the lack of a direct correlation between the linearity of a program and that of its termination arguments.

The program in Figure~\ref{fig:motivation.b} breaks a different assumption made by the majority of termination analyses, namely that \emph{ranking functions are linear}.
Our technique proves that this loop terminates by finding that $|x|$ is a possible ranking function for it.

Figure~\ref{fig:motivation.c} invalidates the assumption that \emph{integer semantics and bit-vector semantics are equivalent}. 
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.

The same assumption is broken by the program in Figure~\ref{fig:motivation.d}. Every execution of the program can be partitioned into two phases: in the first phase $y$ increases
until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
This example is taken from \cite{DBLP:conf/tacas/LeikeH14}, where the authors make use of a template for obtaining a ranking function that proceeds
through a fixed number of phases in the program execution. Each phase is ranked by a linear function, and ends when this function becomes non-positive.
However, when considering the bit-vector semantics, this program does not terminate as..

The loops in Figure~\ref{fig:motivation.e} and Figure~\ref{fig:motivation.f} break the assumption that \emph{real semantics and float semantics are equivalent}. 
% ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
This assumption may lead to erroneous diagnosis of a program's terminating behaviour.% as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
While the former does not terminate for reals, but does for floats, the latter 
terminates for reals, but does not for floats.

Figure~\ref{fig:motivation.g} is a linear program taken from \cite{DBLP:conf/tacas/CookSZ13}, where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a disjunctively well-founded one, which requires an expensive binary reachability analysis.
However, by using our technique we can find a non-linear lexicographic ranking functions.
A lexicographic non-linear ranking functions consists of lexicographically ordered components
of non-linear functions. As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
loop transition leads to a decrease with respect to the lexicographic
ordering for this tuple. Therefore no function may increase unless a function of
a lower index decreases. Additionally, at every step, there must be at least one
function that decreases.

The program in Figure~\ref{fig:motivation.h} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multi-phase ranking function, i.e. ..,  with a dedicated technique for computing it. However, in our setting 
this type of programs do not need a special treatment, as we can find a non-linear ranking function for it as follows:
\begin{verbatim}
R(x, y, z) = (x < y, z)
\end{verbatim}

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}[language=C]
 while(x > 0) x++;
 \end{lstlisting}
\caption{}
 \label{fig:motivation.c}
\end{subfigure} \\

\hline

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
  while ( q >= 0 ) {
    q = q - y;
    y = y + 1;
  }
\end{lstlisting}
\caption{}
 \label{fig:motivation.d}
\end{subfigure}

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0.0) x -= 1.0;
\end{lstlisting}
\caption{}
 \label{fig:motivation.e}
\end{subfigure}

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0.0) x *= 0.5;
\end{lstlisting}
\caption{}
 \label{fig:motivation.f}
\end{subfigure} \\
\hline

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
  while (x != 0) {
    if (x > 0)
      x--;
    else
      x++;
  }
\end{lstlisting}
\caption{}
 \label{fig:motivation.g}
\end{subfigure}

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0 && y > 0 && z > 0) {
    if (y > x) {
      y = z;
      x = nondet();
      z = x - 1;
    } else {
      z = z - 1;
      x = nondet();
      y = x - 1;
    }
}
\end{lstlisting}
\caption{}
 \label{fig:motivation.h}
\end{subfigure}

\end{tabular}
\caption{Sample programs}\label{fig:motivation}
\end{figure*}

%\subsection{Differences in the termination behaviour for integers and bit-vectors.}
%We have collected a number of motivational examples from other termination papers that treat bit-vectors as mathematical integers \cite{DBLP:conf/tacas/LeikeH14,DBLP:conf/tacas/CookSZ13}. 
%We show that the termination arguments computed by such techniques do not directly apply when considering the bit-vector semantics.
%these programs actually exhibit different terminating behaviours for bit-vectors and integer

%% \begin{lstlisting}
%%   while (x != m) {
%%     if (x > m)
%%       x = 0;
%%     else
%%       x++;
%%   }
%% \end{lstlisting}

%% The program in Figure~\ref{} is taken from \cite{DBLP:conf/tacas/CookSZ13}, where
%% $m$ and $x$ start as any integers with $m$ positive. If $x$ is greater than
%% $m$, $x$ is set to 0. Otherwise, $x$ increases until it equals $m$, upon which
%% the loop terminates. While a disjunctive well-founded termination argument does exist for the loop, 
%% e.g. ($x < oldx$ and $0 \leq oldx$) or ($m-x < oldm-oldx$ and $0 \leq oldm-oldx$), the loop does not 
%% terminate under the bit-vector semantics. The reason is ..

%\subsection{Multi-phase ranking functions}
%\subsection{Lexicographic ranking function with strict inequalities.}
%Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.

@@

As well as machine arithmetic differing from Peano arithmetic on the operators they have in common,
computers have several ``bitwise'' operations that are not taken as primitive in the theory of
integers.  These operations include the standard boolean operations \texttt{and, or, not, xor}
applied to each element of the bit-vector.  Computer programs often make use of these operators,
which are non-linear when interpreted in the standard model of Peano arithmetic
(although some of these operators can be seen as linear in a different algebraic structure,
e.g. \texttt{xor} corresponds to addition in the Galois field $\mathrm{GF}(2^k)$).

@@

\section{Termination as Second-Order Satisfaction}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
we will refer to as the \emph{synthesis fragment}.  This fragment
consists of $\exists \forall$ formulae where the the universally quantified variables only range over
ground terms and has close ties to the program synthesis problem.  We will make use of this observation
to frame the halting problem as a program synthesis problem, allowing us to employ the program synthesiser described in
Section~\ref{sec:synthesis}.

@@

\begin{definition}[Conditional Termination Formula]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, V \cdot \forall x, x' \cdot & I(x) \rightarrow V(x) ~ \wedge \\
                                 & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Multiple Loops Termination Formula]
\label{def:multi-termination-formula}
 \begin{align*}
  \exists R, V_1, V_2 \cdot \forall x, x' \cdot & I(x) \rightarrow V_1(x) ~ \wedge \\
                                        & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
                                        & V_1(x) \wedge \lnot L_1(x) \rightarrow V_2(x) ~ \wedge \\
                                        & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

A slightly more involved case occurs if $L$'s termination depends on the state it begins
executing in, i.e. $L$ is \emph{conditionally terminating}.  This property is equivalent to the
formula of Defintion~\ref{def:conditional-termination-formula}.  If this formula is satisfiable,
$V$ is an inductive invariant of $L$ that is established by the initial states $I$.
$R$ is then a ranking function for $L$ as restricted by $V$ -- that is to say, $R$ need only
be well founded on those states satisfying $V$.  But since $V$ is an inductive invariant of $L$,
$R$ is strong enough to show that $L$ terminates from any of its initial states.  We say
that $V$ is a \emph{termination invariant} for $L$ and that $R$ proves termination relative to $V$.

@@

With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae
belongs to the synthesis fragment.  We use this name because the program synthesis problem can be
described as finding a satisfying assigment to the following formula, which is the general case of
the synthesis fragment:

\begin{definition}[Synthesis Formula]
 \label{def:synth-formula}
 $$\exists P, x \cdot \forall y \cdot \sigma(x, y, P)$$
\end{definition}

Here we can view $\sigma: X \times Y \times (X \times Y \to Z) \to \mathbb{B}$ as a specification function.
The idea is that $\sigma$ returns true iff the function $P$ computes appropriate outputs
when fed the inputs $x$ and $y$.  If we can find a function $P$ and a value $x$ that meet the specification
for all $y$, we have found a solution to the synthesis problem and write $P \models \sigma$.  Since the formulae
of Fig.~\ref{fig:termination-formulae} are syntactically in the synthesis fragment, by concocting an appropriate
specification function we reduce the problem of proving termination to the problem of
program synthesis.

\subsection{Synthesis Complexity}
In Section~\ref{sec:synthesis} we will describe a procedure for solving the synthesis formula.
This procedure has the property that it will find the shortest program meeting the
specification, and as a consequence the runtime of the program synthesiser is dominated
by the size of this shortest program.  To frame the discussion of the runtime of our
termination proving procedure, we first recall the definition of the Kolmogorov complexity
of a function $f$:
